Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks

### 2025-05-12 inst/paper/main.tex
update results on onset-to-isolation scenarios

We assume that an isolated case cannot transmit, so isolation is assumed to be 100\% effective once enacted ($R_{isolated} = 0$). The onset-to-isolation delay is parameterised with three scenarios: a fast response, based on the SARS in Hong Kong (Donnelly et al. 2003) and a slower response based on COVID-19 in Wuhan, China (Li et al. 2020) (Figure \ref{fig:onset-to-isolation}). Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:onset-to-isolation}). These two empirically informed delay distributions provide a guide to possible response times in the case of an influenza outbreak and will enable us to evaluate if such delays are too long to contain a disease that can transmit rapidly such as influenza. We also include a third onset-to-isolation scenario based on the case where lateral flow rapid tests (LFTs) are provided to all individuals in the susceptible population and these people are freely able to self-test if they suspect they are symptomatic. We parameterise this as an exponential distribution, with rate ($\lambda$) 0.5, where most individuals will self-test very shortly after they become symptomatic (approximately 40\% of cases will self-test within the first day of sympmtom onset), with a few people waiting a 2 or more days before testing (approximately 3\% take longer than one week to take a self-test). \\

@@

The incubation period for H1N1 is noticable shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:incub}). \\

Predictably, the ability to control an influenza epidemic is highly dependent on the reproduction number. A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of control effectiveness. (Figure \ref{fig:prop-outbreak-control-R}). There is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of isolation and contact tracing for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}). As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e. 0\% control effectiveness), however, outbreaks with $R = 1.5$ are mostly contained once control effectiveness exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}. Once the reproduction number is at 2.5 or 3.5, an outbreak is only controlled in at least 50\% of scenarios with isolation and contact tracing at greater than 60\% (Figure \ref{fig:prop-outbreak-control-R}). \\

The incubation period influences the efficacy of isolating and tracing. This effect is most pronounced when the percentage of outbreaks controlled is greater than 0\% and less than 100\% (i.e. not at the boundary) (Figure \ref{fig:prop-outbreak-control-R}). The shortest incubation period, H1N1, predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection limit our ability to effectively tracing and isolate their potential infectees in time to prevent disease transmission. The difficulty to control an pandemic outbreak with a short incubation period is shown by the $\sim 40\%$ of simulated H1N1 outbreaks with 100\% control effectiveness that were not contained when $R = 3.5$, while longer incubation period outbreaks have $>80\%$ controlled when 100\% of contacts are traced and isolated (Figure \ref{fig:prop-outbreak-control-R}). Overall, both the transmissibility and incubation period impact the feasibility to control an outbreak. \\

The variance between the influenza subtypes at each percentage of contacts traced and isolated varies (i.e. difference between circle, triangle and square points at each contacts traced \% in Figure \ref{fig:prop-outbreak-control-R}). This variance is greatest at intermediate (20\% - 80\%) contact tracing effectiveness and intermediate values of $R$ (1.5 and 2.5) (Figure \ref{fig:prop-outbreak-control-R}). This could indicate a substantive role for incubation period durartion in the control of disease transmission. However, this variance in proportion of outbreaks controlled between pathogen subtypes does not show a strong pattern across all scenarios simulated (Figure \ref{fig:prop-outbreak-control-var-R}). There is a weak pattern of increased variance in controllability between subtypes at intermediate values of $R$, overall it seems that incubation period causes differences in control-by-isolation effectivness when not at the boundary of epidemic controllability. In other words, differences between influenza subtypes arise when not in a situation of high transmissibility and low contact tracing effectivness that cannot contain any outbreaks, nor low transmissibility and high contact tracing effectivenss easily contain all outbreaks (Figure \ref{fig:prop-outbreak-control-var-R}). \\

>>>>

We assume that an isolated case cannot transmit, so isolation is assumed to be 100\% effective once enacted ($R_{isolated} = 0$). The onset-to-isolation delay is parameterised with three scenarios: a fast response, based on the SARS in Hong Kong (Donnelly et al. 2003) and a slower response based on COVID-19 in Wuhan, China (Li et al. 2020) (Figure \ref{fig:onset-to-isolation}). Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:onset-to-isolation}). These two empirically informed delay distributions provide a guide to possible response times in the case of an influenza outbreak and will enable us to evaluate if such delays are too long to contain a disease that can transmit rapidly such as influenza. We also include a third onset-to-isolation scenario based on the case where lateral flow rapid tests (LFTs) are provided to all individuals in the susceptible population and these people are freely able to self-test if they suspect they are symptomatic. We parameterise this as an exponential distribution, with rate ($\lambda$) 0.5, where most individuals will self-test very shortly after they become symptomatic (approximately 40\% of cases will self-test within the first day of sympmtom onset), with a few people waiting a 2 or more days before testing (approximately 3\% take longer than one week to take a self-test). We assume that LFTs are 100\% accurate so all symptomatic individuals will receive a positive test result. \\

@@

The incubation period for H1N1 is noticably shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:incub}). \\

Predictably, the ability to control an influenza epidemic is highly dependent on the reproduction number and the speed of isolation after symptom onset (Figure \ref{fig:prop-outbreak-control-R}). A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of contacts traced, except in cases where they are equal because zero or all outbreaks are contained, across all onset-to-isolation scenarios (Figure \ref{fig:prop-outbreak-control-R}). When the reponse to isolate cases is slow then outbreaks with $R = 1.5$ require at least 50\% of contacts traced to contain at least half of simulated outbreaks. When $R$ is 2.5 or 3.5 then contact tracing needs to be extremely comprehensive ($>90\%$ to have a chance at containment, and even then $R = 3.5$ influenza outbreaks on average are uncontrolled (Figure \ref{fig:prop-outbreak-control-R}A). Especially for H1N1, given it's faster tempo of transmission, having a much lower probability of containment at 100\% of contacts traced with a slow, COVID-19-like, isolation response time (Figure \ref{fig:prop-outbreak-control-R}A). In a scenario with a fast, Hong Kong SARS-like, onset-to-isolation response, outbreaks with $R = 1.1$ are almost always contained at any proportion of contacts traced (Figure \ref{fig:prop-outbreak-control-R}B). For slow and fast isolation response delays there is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of isolation and contact tracing for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}A,B). As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e. 0\% control effectiveness), however, outbreaks with $R = 1.5$ are mostly contained once percentage of contacts ascertained in tracing exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}B). For more transmissible influenza scenarios, even a fast isolation response with an average of 3.4 days between symptom onset and isolation requires $>70\%$ of contacts traced for most outbreaks to be controlled (Figure \ref{fig:prop-outbreak-control-R}B). \\

The most rapid scenario we simulated to emulate individuals self-testing with LFTs, has the same trend of increased chance of outbreak containment by reducing the time between symptom onset and isolation, as the slow and fast scenarios (Figure \ref{fig:prop-outbreak-control-R}). The contol of low transmissible ($R \in {1.1, 1.5})$ influenza pathogens is almost always guranteed even in the absence of contact tracing and only isolating cases immediately if positive on an LFT (Figure \ref{fig:prop-outbreak-control-R}C). Assuming ubiquitous access to home testing and that 63\% of symptomatic cases testing and isolating within the 24 hours symptom onset, 75\% ascertainment of contacts results in most outbreaks controlled for all influenza transmissibility scenarios ($R \leq 3.5$). The LFT isolation strategy is also the only onset-to-isolation delay which can achieve 100\% of outbreaks controlled when all contacts are traced (Figure \ref{fig:prop-outbreak-control-R}C). \\

The difference in influenza subtype incubation period influences the efficacy of isolating and tracing. This effect is most pronounced when the percentage of outbreaks controlled is greater than 0\% and less than 100\% (i.e. not at the boundary) (Figure \ref{fig:prop-outbreak-control-R}). The shortest incubation period, H1N1, predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection and in some instances transmission while subclinical limit our ability to effectively tracing and isolate their potential infectees in time to prevent disease transmission. The variance between the influenza subtypes at each percentage of contacts traced and isolated varies (i.e. difference between circle, triangle and square points at each contacts traced \% in Figure \ref{fig:prop-outbreak-control-R}). This variance is greatest at intermediate (20\% - 80\%) contact tracing effectiveness and intermediate values of $R$ (1.5 and 2.5) (Figure \ref{fig:prop-outbreak-control-R}). This could indicate a substantive role for incubation period durartion in the control of disease transmission. However, this variance in proportion of outbreaks controlled between pathogen subtypes does not show a strong pattern across all scenarios simulated (Figure \ref{fig:prop-outbreak-control-var-R}). There is a weak pattern of increased variance in controllability between subtypes at intermediate values of $R$, overall it seems that incubation period causes differences in control-by-isolation effectivness when not at the boundary of epidemic controllability. In other words, differences between influenza subtypes arise when not in a situation of high transmissibility and low contact tracing effectivness that cannot contain any outbreaks, nor low transmissibility and high contact tracing effectivenss easily contain all outbreaks (Figure \ref{fig:prop-outbreak-control-var-R}). \\
